The dynamics of starvation and recovery

### 2016-07-26 Manuscript/3.tex
sid updates

Whereas the relation between consumer growth rate $\lambda$ and the
starvation rate $\sigma$ defines an absolute bound of biological
feasibility---the TC bifurcation---the starvation rate $\sigma$ also
determines the sensitivity of the consumer population to changes in resource
density.  When $\sigma\gg\lambda$, the steady-state population density is
small, thereby increasing the risk of stochastic extinction.  On the other
hand, as $\sigma$ decreases, the system will ultimately be poised either near
the TC or the Hopf bifurcation (Fig.~\ref{fig:fp}).  If the recovery rate
$\rho$ is sufficiently small, the TC bifurcation is reached and the resource
eventually is eliminated.  If $\rho$ exceeds a threshold value, cyclic
dynamics will develop as the Hopf bifurcation is approached.
\\

@@

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features~\cite{}\sid{Ref}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $M$ is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans~\cite{}\sid{Ref}, and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. Several efforts have shown how a
partitioning of this metabolic rate between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{Kempes:2012hy,more}\sid{fix Ref}. More specifically, the interspecific
\sid{what does this word mean?}  trends in growth rate can be approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development. It is useful to explicitly write this balance
because it can also be modified to understand the timescales of both
starvation and recovery from starvation as we show below.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from its replete \sid{replete vs. adult} state
to a reduced-size hungry state where reproduction is impossible.  This
transition time can be inferred from the energy balance in Eq.~\ref{balance},
where we make the basic assumption that an organism must meet its maintenance
requirements using the digestion of existing mass as the sole energy source.
This assumption implies the simple metabolic balance \sid{don't understand
  this equation}
\begin{eqnarray}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$~\cite{}\sid{Ref}, the energy required to
synthesis a unit of biomass. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-B_{m}t/E_{m}^{\prime}}.
\end{eqnarray}
Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation and
death by the timescales required to reach specific fractions of normal adult
mass. We define $m_{\rm starve}=\epsilon M$, where $\epsilon<1$ is the
fraction of adult mass where reproduction ceases.  This fraction will be
modified if tissue composition systematically scales with adult mass.  For
example, making use of the observation that body fat in mammals scales with
overall body size according to $M_{f}=f_{0}M^{\gamma}$ and assuming that once
this mass is fully digested the organism begins to starve, this would imply
that $\epsilon=1-f_{0}M^{\gamma}/M$.  Using this criterion in Eq.~\ref{mt},
the time scale for starvation is given by
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-\frac{E_{m}\ln\left(\epsilon\right)}{B_{m}}.
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with replete
mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important feature is
that $\sigma$ does not have a simple scaling dependence on $\lambda$
(Eq.~\ref{lambda}), which is important for the dynamics that we later
discuss.

The time to death should follow a similar relation, but defined by a lower
fraction of adult mass, $m_{\rm death}=\epsilon^{\prime} M$.  Suppose, for
example, that an organism dies once it has digested all fat and muscle
tissues, and that muscle tissue scales with body mass according to
$M_{mm}=mm_{0}M^{\zeta}$ \sid{what is $mm_0$?}.  This gives
$\epsilon^{\prime}=1-\left(f_{0}M^{\gamma}+mm_{0}M^{\zeta}\right)/M$. Muscle
mass has been shown to be roughly proportional to body mass~\cite{muscle} in
mammals and thus $\epsilon^{\prime}$ is merely $\epsilon$ minus a constant.
Thus
\begin{eqnarray}
t_{\mu}=-\frac{E_{m}\ln\left(\epsilon^{\prime}\right)}{B_{m}}
\end{eqnarray}
and $\mu=1/t_{\mu}$.

@@

The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the starving \sid{starving vs.\ hungry}
state to the full state.  We again use the balance given in Eq.~\ref{balance}
to find the timescale to return to the replete mass from a given reduced
starvation mass.  From the solution to Eq.~\ref{balance}
\begin{eqnarray}
m\left(t\right)\!=\!\left(\!\frac{B_{0}}{B_{m}}\!\right)^{1/(\eta-1)}\!\!
\left[1\!-\!\left(1\!-\!\frac{B_{m}}{B_{0}}m_{0}^{1\!-\!\eta}\right)e^{-b\left(1\!-\!\eta\right)t}\right]^{1/\left(1-\eta\right)}%\left(\frac{a}{b}\right)^{1/\left(1-\eta\right)}\,,
\end{eqnarray}
we require the timescale, $t_{\rho}=t_{2}-t_{1}$, which is the time it takes
to go from $m\left(t_{1}\right)=\epsilon M$ to $m\left(t_{2}\right)=M$, or
\begin{eqnarray}
t_{\rho}=\frac{E_{m}\left\{\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(M
  \right)^{1\!-\!\eta }\right]
-\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(\epsilon M \right)^{1\!-\!\eta }\right]\right\}}{(\eta -1) B_{m}}.
\end{eqnarray}
%\left(\frac{a}{b}\right)^{\frac{1}{\eta -1}}
%where $c=(a/b)^{1/(\eta-1)}$.
Although these rate equations are general, here we focus on parameterizations
for terrestrial-bound endotherms, specifically mammals, which range from a
minimum of $M\approx1$ gram (the Etruscan shrew \emph{Suncus etruscus}) to a
maximum of $M\approx10^7$ grams (the late Eocene to early Miocene
Indricotheriinae).  Investigating other classes of organisms would simply
involve altering the metabolic exponents and scalings associate with
$\epsilon$. Moreover, we emphasize that our allometric equations describe
mean relationships, and do not account for the (sometimes considerable)
variance associated with individual species.
\\
%, though this is beyond the current scope of our analysis. %, and we henceforth focus our examination on mammalian species.

@@

\noindent {\bf Extinction risk}\\
Within our model, higher rates of starvation result in a larger flux of the
population to the hungry state.  In this state reproduction is absent, thus
increasing the likelihood of extinction.  However, from the perspective of
population survival, it is the rate of starvation relative to the rate of
recovery that determines the long-term dynamics of the system
(Fig.~\ref{fig:fp}).  We now examine the competing effects of cyclic dynamics
vs.\ changes in steady state density on extinction risk as a function of the
ratio $\sigma/\rho$.  To this end, we computed the probability of extinction,
where extinction is defined as the population trajectory going below
$0.2~\times$ the allometrically constrained steady state for all times
between $10^2$ and $\leq 10^6$.  This procedure is repeated for 1000
replicates \sid{what is changing? isn't the system deterministic?} of the
continuous-time system shown in Eq.~\ref{eq:system} for an organism of
$M=100$ grams, assuming random initial conditions around the steady state
(Eq.~\ref{eq:ss}). %random initial conditions within the interval $(R^*,H^*,F^*)\pm100\%$.
By allowing the rate of starvation to vary, we assessed extinction risk
across a range of values of the ratio $\sigma/\rho$ varying between $10^{-2}$
to $2.5$, thus examining a horizontal cross-section of Fig.~\ref{fig:fp}.  As
expected, higher rates of extinction correlated with both low and high values
of $\sigma/\rho$.  For low values of $\sigma/\rho$, the increased extinction
risk results from transient cycles with larger amplitudes as the system nears
the Hopf bifurcation (Fig.~\ref{fig:ext}).  For large values of
$\sigma/\rho$, higher extinction risk arises because of to the decrease in
the steady state consumer population density.  This interplay creates an
`extinction refuge' as shown in Fig.~\ref{fig:ext}, such that for a
relatively constrained range of $\sigma/\rho$, extinction probabilities are
minimized.

>>>>

Whereas the relation between consumer growth rate $\lambda$ and the
starvation rate $\sigma$ defines an absolute bound of biological
feasibility---the TC bifurcation---the starvation rate $\sigma$ also
determines the sensitivity of the consumer population to changes in resource
density.  When $\sigma\gg\lambda$, the steady-state population density is
small, thereby increasing the risk of stochastic extinction.  On the other
hand, as $\sigma$ decreases, the system will ultimately be poised either near
the TC or the Hopf bifurcation (Fig.~\ref{fig:fp}).  If the recovery rate
$\rho$ is sufficiently small, the TC bifurcation is reached and the resource
eventually is eliminated.  If $\rho$ exceeds a threshold value, cyclic
dynamics will develop as the Hopf bifurcation is approached.

\sid{mention  and refer to the starving random walk model somewhere.}
\\

@@

Nearly all of the rates described in the NSM are to some extent governed by
consumer metabolism, which can be used to describe a variety of organismal
features~\cite{}\sid{Ref}.  The scaling relation between an organism's
metabolic rate $B$ and its body size at reproductive maturity $M$ is well
documented~\cite{West:2002it} and scales as $B = B_0 M^\eta$, where $\eta$ is
the scaling exponent, generally assumed to vary around $2/3$ or $3/4$ for
metazoans~\cite{}\sid{Ref}, and has taxonomic shifts for unicellular species
between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}. Several efforts have shown how a
partitioning of this metabolic rate between growth and maintenance purposes
can be used to derive a general equation for the growth trajectories and
growth rates of organisms ranging from bacteria to metazoans
\cite{Kempes:2012hy,more}\sid{fix Ref}. More specifically, the interspecific
\sid{what does interspecific mean?}  trends in growth rate can be
approximated by
\begin{eqnarray}
\lambda = \lambda_0 M^{\eta-1}.
\label{lambda}
\end{eqnarray}
This relation is derived from the simple balance condition
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
at any point in development. It is useful to explicitly write this balance
because it can also be modified to understand the timescales of both
starvation and recovery from starvation as we show below.

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$ (henceforth we term this state as the ``replete'' state), to a
reduced-size hungry state where reproduction is impossible.  This transition
time can be inferred from the energy balance in Eq.~\ref{balance}, where we
make the basic assumption that an organism must meet its maintenance
requirements using the digestion of existing mass as the sole energy source.
This assumption implies the simple metabolic balance \sid{don't understand
  this equation}
\begin{eqnarray}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$~\cite{}\sid{Ref}, the energy required to
synthesis a unit of biomass. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-B_{m}t/E_{m}^{\prime}}.
\end{eqnarray}
Since only certain tissues can be digested for energy (for example the brain
cannot be degraded to fuel metabolism), we define the rate for starvation and
death by the timescales required to reach specific fractions of the
replete-state mass.  We define $m_{\rm starve}=\epsilon M$, where
$\epsilon<1$ is the fraction of replete-state mass where reproduction ceases.
This fraction will be modified if tissue composition systematically scales
with adult mass.  For example, making use of the observation that body fat in
mammals scales with overall body size according to $M_{f}=f_{0}M^{\gamma}$
and assuming that once this mass is fully digested the organism begins to
starve, this would imply that $\epsilon=1-f_{0}M^{\gamma}/M$.  Using this
criterion in Eq.~\ref{mt}, the time scale for starvation is given by
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-\frac{E_{m}\ln\left(\epsilon\right)}{B_{m}}.
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Eq.~\ref{lambda}), which is important for the dynamics that we
later discuss.

The time to death should follow a similar relation, but defined by a lower
fraction of replete-state mass, $m_{\rm death}=\epsilon^{\prime} M$.
Suppose, for example, that an organism dies once it has digested all fat and
muscle tissues, and that muscle tissue scales with body mass according to
$M_{mm}=mm_{0}M^{\zeta}$ \sid{what is $mm_0$?}.  This gives
$\epsilon^{\prime}=1-\left(f_{0}M^{\gamma}+mm_{0}M^{\zeta}\right)/M$. Muscle
mass has been shown to be roughly proportional to body mass~\cite{muscle} in
mammals and thus $\epsilon^{\prime}$ is merely $\epsilon$ minus a constant.
Thus
\begin{eqnarray}
t_{\mu}=-\frac{E_{m}\ln\left(\epsilon^{\prime}\right)}{B_{m}}
\end{eqnarray}
and $\mu=1/t_{\mu}$.

@@

The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the starving \sid{starving vs.\ hungry}
state to the full state.  We again use the balance given in Eq.~\ref{balance}
to find the timescale to return to the replete-state mass from a given
reduced starvation mass.  From the solution to Eq.~\ref{balance}
\begin{eqnarray}
m\left(t\right)\!=\!\left(\!\frac{B_{0}}{B_{m}}\!\right)^{1/(\eta-1)}\!\!
\left[1\!-\!\left(1\!-\!\frac{B_{m}}{B_{0}}m_{0}^{1\!-\!\eta}\right)e^{-b\left(1\!-\!\eta\right)t}\right]^{1/\left(1-\eta\right)}%\left(\frac{a}{b}\right)^{1/\left(1-\eta\right)}\,,
\end{eqnarray}
we require the timescale, $t_{\rho}=t_{2}-t_{1}$, which is the time it takes
to go from $m\left(t_{1}\right)=\epsilon M$ to $m\left(t_{2}\right)=M$, or
\begin{eqnarray}
t_{\rho}=\frac{E_{m}\left\{\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(M
  \right)^{1\!-\!\eta }\right]
-\ln \left[1\!-\!\frac{B_{0}}{B_{m}}\left(\epsilon M \right)^{1\!-\!\eta }\right]\right\}}{(\eta -1) B_{m}}.
\end{eqnarray}
%\left(\frac{a}{b}\right)^{\frac{1}{\eta -1}}
%where $c=(a/b)^{1/(\eta-1)}$.
Although these rate equations are general, here we focus on parameterizations
for terrestrial-bound endotherms, specifically mammals, which range from a
minimum of $M\approx1$ gram (the Etruscan shrew \emph{Suncus etruscus}) to a
maximum of $M\approx10^7$ grams (the late Eocene to early Miocene
Indricotheriinae).  Investigating other classes of organisms would simply
involve altering the metabolic exponents and scalings associate with
$\epsilon$. Moreover, we emphasize that our allometric equations describe
mean relationships, and do not account for the (sometimes considerable)
variance associated with individual species.
\\
%, though this is beyond the current scope of our analysis. %, and we henceforth focus our examination on mammalian species.

@@

\noindent {\bf Extinction risk}\\
Within our model, higher rates of starvation result in a larger flux of the
population to the hungry state.  In this state reproduction is absent, thus
increasing the likelihood of extinction.  However, from the perspective of
population survival, it is the rate of starvation relative to the rate of
recovery that determines the long-term dynamics of the system
(Fig.~\ref{fig:fp}).  We now examine the competing effects of cyclic dynamics
vs.\ changes in steady state density on extinction risk as a function of the
ratio $\sigma/\rho$.  To this end, we computed the probability of extinction,
where extinction is defined as the population trajectory going below
$0.2~\times$ the allometrically constrained steady state for all times
between $10^2$ and $\leq 10^6$.  This procedure is repeated for 1000
replicates of the continuous-time system shown in Eq.~\ref{eq:system} for an
organism of $M=100$ grams.  In each replicate the initial condition is
distributed around the steady state (Eq.~\ref{eq:ss}).  Specifically the
initial densities are chosen to be $A(F^*,H^*,R^*)$, with $A$ a random
variable that is uniformly distribution in $[0,2]$.  By allowing the rate of
starvation to vary, we assessed extinction risk across a range of values of
the ratio $\sigma/\rho$ varying between $10^{-2}$ to $2.5$, thus examining a
horizontal cross-section of Fig.~\ref{fig:fp}.  As expected, higher rates of
extinction correlated with both low and high values of $\sigma/\rho$.  For
low values of $\sigma/\rho$, the increased extinction risk results from
transient cycles with larger amplitudes as the system nears the Hopf
bifurcation (Fig.~\ref{fig:ext}).  For large values of $\sigma/\rho$, higher
extinction risk arises because of to the decrease in the steady state
consumer population density.  This interplay creates an `extinction refuge'
as shown in Fig.~\ref{fig:ext}, such that for a relatively constrained range
of $\sigma/\rho$, extinction probabilities are minimized.
